---
title: "Design-based and model-based inference and spatial sampling"
author: "Edzer Pebesma, Hanna Meyer"
date: "EO4Statistics meeting, Nov 12 2020"
output:
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1331)
```

## References

Most of what follows is based on the following references:

* De Gruijter, J. J., & Ter Braak, C. J. F. (1990). Model-free estimation from spatial samples: a reappraisal of classical sampling theory. Mathematical geology, 22(4), 407-415.
* Brus, D. J., & De Gruijter, J. J. (1997). Random sampling or geostatistical modelling? Choosing between design-based and model-based sampling strategies for soil (with discussion). Geoderma, 80(1-2), 1-44.
* De Gruijter, J., Brus, D. J., Bierkens, M. F., & Knotters, M. (2006). Sampling for natural resource monitoring. Springer Science & Business Media.

--------

```{r echo=FALSE}
suppressPackageStartupMessages(library(gstat))
suppressPackageStartupMessages(library(stars))
grd = st_as_stars(expand.grid(x = 1:100, y = 1:100))
grd$x = 1
v <- vgm(1, "Sph", 40)
x <- krige(x ~ 1, locations = NULL, newdata = grd, model = v, 
	nmax = 20, beta = 0, nsim = 6, dummy = TRUE, debug.level = 0)
x$f = factor((x < 0.5)[[1]], labels = c("Forest", "Non-Forest"))
plot(x["f",,,1], reset = FALSE, key.pos = 4, key.length = .4, key.width = lcm(5), main = NULL)
aoi = st_as_sfc(st_bbox(x))
plot(aoi, col = NA, border = 'red', add = TRUE, lwd = 2)
```

What is the proportion of $p_F$ of forest in the area of interest (box)?

* Given we can plan the sampling, how should we do that?
* Given a sample has already been taken, what are our options to estimate $p_F$?

-------

```{r echo=FALSE}
n = 50
s = st_sample(aoi, n)
plot(x["f",,,1], reset = FALSE, key.pos = 4, key.length = .4, key.width = lcm(5), main = NULL)
plot(s, add = TRUE, col = 'green', pch = 3)
v = st_extract(x["f",,,1], s)
```

## 95% CI, design-based, simple random sampling

Sample mean:
```{r}
(m = mean(v$f == "Forest"))
```

95% CI using normal approximation (CLT, n large, $m$ not too close to 0 or 1):
```{r}
n
se = sqrt(m * (1 - m) / n)
m + qnorm(c(.025,.975)) * se
```

## What does this confidence interval mean?

(we use small $s$ for a fixed variable, capital $S$ for a random variable)

* it is a design-based estimate: observations on a variable $z$ at locations $s$ are considered as coming from a random sampling process, $z(S)$, with $S$ randomly sampled from the area of interest $A$: $S$ is a random variable, $z$ is fixed (as is $z(s)$).
* because of random sampling, _any_ pair $\{z(S_i),~z(S_j)\}$ are independent, and their separation distance $|S_i-S_j|$ is again a random variable
* the CI means that **under repeated sampling** (using the same procedure), on average 95 out of 100 cases this CI will include the population parameter $p_F$.

## Replicates: repeated sampling

```{r,echo=FALSE,fig.width=9.5,fig.height=2.5}
nr = 6
layout(matrix(1:11, 1), widths = c(1,.1,1,.1,1,.1,1,.1,1,.1,1))
for (i in 1:nr) {
  plot(x["f",,,1], reset = FALSE, key.pos = NULL, main = NULL)
  plot(st_sample(aoi, n), add = TRUE, col = 'green', pch = 3)
  if (i < nr) plot.new()
}
```
...

## CI coverage

```{r,echo=FALSE,fig.width=9.5,fig.height=5}
m = sapply(1:100, function(y) mean(st_extract(x["f",,,1], st_sample(aoi, n))$f == "Forest", 
	na.rm = TRUE))
se = sqrt(m * (1 - m) / n)
int = cbind(m + qnorm(.025) * se, m + qnorm(.975) * se)
plot(1:100, m, pch = 3, cex = .5, xlim = c(0,101), ylim = c(0, .7), xlab = "sample", ylab = "fraction forest")
segments(1:100, int[,1], y1 = int[,2])
abline(h = mean(m), lty = 2, col = 'red')
```

(red line: population mean)

## 95% CI, model-based, whatever sampling strategy

* model-based, we assume $z(s)$ to be a realisation (sample of size 1) of a superpopulation (random field) $Z(s)$
* we postulate (assume) a model, e.g. $Z(s) = m + e(s)$, with $m$ a spatial constant unknown mean (fixed effect), and $e(s)$ a spatially correlated random variable (random effect)
* For stationary random fields, Correlation of $Z(s_i)$ and $Z(s_j)$ are assumed to be correlated as a function of distance: $\mbox{Cov}(Z(s_i), Z(s_j)) = C(h)$, with $h=|s_i-s_j|$
* sampling strategy does "not" play a role

## Estimating $m$?

```{r,echo=FALSE,fig.width=9.5,fig.height=2.5}
nr = 6
layout(matrix(1:11, 1), widths = c(1,.1,1,.1,1,.1,1,.1,1,.1,1))
for (i in 1:nr) {
  plot(x["f",,,i], reset = FALSE, key.pos = NULL, main = NULL)
  plot(s, add = TRUE, col = 'green', pch = 3)
  if (i < nr) plot.new()
}
```

* $m$ is a property of the whole random field $Z(s)$, not just the realisation at hand
(unconditional to the observations)
* it is also in no way spatially restricted to the area of interest

## Predicting $Z(A)$

* Model-based, $s$ is no longer random so we predict $Z(s_0)$, by $\hat{Z}(s_0)=\sum_{i=1}^n \lambda_i Z(s_i)$ with $\lambda_i$ the kriging weights
* This means that we can also predict the area mean $Z(A)=\frac{1}{|A|}\int_A Z(u)du$, e.g.
using the mean of predictions over the area: 
$$\frac{1}{K}\sum_{j=1}^K \hat{Z}(s_j)$$ 
this is the _block kriging_ prediction; it comes with a block
kriging standard error that takes all the correlations between
the $\hat{Z}(s_j)$ into account: block kriging deals with _change
of support_

 
## Required: variogram

```{r echo=FALSE}
vg = variogram(f == "Forest" ~1, na.omit(v))
v.fit = fit.variogram(vg, vgm(1, "Sph", 30))
plot(vg, v.fit)
```

## Kriging predictions:

```{r echo=FALSE,fig.width=10}
kr = krige(f == "Forest" ~ 1, v, grd, v.fit, debug.level = 0)
kr$Population = x["f",,,1] == "Forest"
names(kr)[1] = "Kriging prediction"
hook = function(x) plot(s, add = TRUE, col = 'green', pch = 3)
plot(merge(kr[c(1,3)]), hook = hook, breaks = "equal")
```

## Block kriging for the whole of $A$

```{r echo=FALSE}
bl = krige(f == "Forest" ~ 1, v, aoi, v.fit, debug.level = 0)
```
```{r echo=FALSE}
c(mean = bl$var1.pred)
ci = bl$var1.pred + qnorm(c(.025,.975)) * sqrt(bl$var1.var)
setNames(ci, c("p.025", "p.975"))
```

Compare to design-based:
```{r echo=FALSE}
c(mean = m <- mean(v$f == "Forest"))
se = sqrt(m * (1 - m) / n)
setNames(m + qnorm(c(.025,.975)) * se, c("p.025", "p.975"))
```

## What does the block kriging CI refer to?

```{r,echo=FALSE,fig.width=9.5,fig.height=2.1}
cs <- krige(f == "Forest" ~ 1, v, grd, v.fit, 
	nmax = 20, beta = .27, nsim = 6, debug.level = 0, indicators = TRUE)
cs$f = factor((cs < 0.5)[[1]], labels = c("Forest", "Non-Forest"))
plot(cs["f"], hook = hook, key.pos = 1, key.length = .4, key.width=lcm(1))
```

Averaging over the _conditional_ random fields $Z(s)|z(s_i),i=1.,,,.n$ that:

* have mean $m$ and covariance $C(h)$
* obey the sample data, i.e. $Z(s_i)=z(s_i)$ for all observations $i$

## Comparison design-based / model-based

1. design-based methods do not need a model, and are more robust: model-based approaches lean on an assumed model.
2. design-based models are robust only if random sampling was used, and inclusion probabilities (sample weights) are known
3. model-based approaches make no assumption about sampling, but also have no guard against biased sampling
4. model-based approaches can make predictions for _any_ location, and predict means over regions of _any_ shape; design-based only estimate _averages_ over the sampled region(s).
5. the probability statements differ: design-based CI's refer to repeating the sampling procedure, model-based CI's refer to (conditional) super-populations under a fixed sample.

## The use of R.

* R is a data science language
* All numbers and figure in these slides were generated by R (where
the commands were mostly suppressed, to avoid clutter): 
    * this makes them 100% reproducible, and
	* a source for further experimentation
	* source at https://github.com/edzer/eo4statistics/blob/master/slides.Rmd
	* literate programming helps explaining code, computation, and intention
* R has a number (> 16K) of extension packages, many of them for spatial statistics
* R has the advantage (over e.g. Python) that installation of extension packages on all kind of computers (running OSX, Windows, or Linux) is relatively painless
